---
title: "BumbleDataEDA"
author: "Zhiling Huang"
date: "2024-12-08"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Cleaning

```{r Data Cleaning}
library(dplyr)
library(ggplot2)

data_bombclimate <- read.csv("data_bombclimate.csv")

# Filter for the continent NAM
data_NAM <- data_bombclimate[data_bombclimate$continent == "NAM", ]
data_NAM <- na.omit(data_NAM)

# Calculate the mean temperature
data_NAM$meanTemp <- (data_NAM$maxPeriodAnnualMeanT + data_NAM$minPeriodAnnualMeanT) / 2

final_data <- data_NAM[, c("species", "kmNorthEquator", "elevation",
                           "periodFrom", "periodTo","meanTemp",
                           "maxPeriodAnnualMeanT", "minPeriodAnnualMeanT")]

head(final_data)
## Sample Size ~ 86673
summary(final_data)
```



### Plot 1: Migration patterns 

```{r p1}
# Load necessary libraries
library(ggplot2)
library(dplyr)

# Read the data
data <- read.csv("final_data_1.csv")

# Prepare data for migration analysis: calculate average latitude per species over each period
migration_data <- data %>%
  group_by(species, periodFrom) %>%
  summarize(Average_Latitude = mean(kmNorthEquator, na.rm = TRUE)) %>%
  ungroup()

# Select top 5 most common species
top_species <- data %>%
  count(species, sort = TRUE) %>%
  top_n(10) %>%
  pull(species)

# Filter data to only include top species
filtered_migration_data <- migration_data %>%
  filter(species %in% top_species)

# Plot migration patterns for the selected species
ggplot(filtered_migration_data, aes(x = periodFrom, y = Average_Latitude, color = species)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(title = "Migration Patterns of Top Bumblebee Species Over Time",
       x = "Starting Year of Period", y = "Average Latitude (km from Equator)",
       color = "Species") +
  theme_minimal() +
  theme(legend.position = "right")

```
### Plot 2: Correlation Plot

```{r p5}
library(ggplot2)
library(GGally)
data$temperature_variation <- data$maxPeriodAnnualMeanT - data$minPeriodAnnualMeanT
ggpairs(data[, c("kmNorthEquator", "temperature_variation")],
        title = "Pairwise Relationships Among Key Variables")
```

## Plot 3: Change in elevation of 31 bumblebee species by 1999 to 2010 relative to their mean latitude

```{r mean elevation}
library(ggplot2)
library(dplyr)
library(gridExtra)

data <- read.csv("final_data_1.csv")

time_intervals <- unique(data$periodTo)

species_summary <- data %>%
  group_by(species, periodTo) %>%
  summarize(
    mean_elevation = mean(elevation, na.rm = TRUE),
    mean_latitude = mean(kmNorthEquator, na.rm = TRUE)
  ) %>%
  ungroup()

plots <- list()
for (interval in time_intervals) {
  # Filter data for the current time interval
  interval_data <- species_summary %>%
    filter(periodTo == interval)
  
  p <- ggplot(interval_data, aes(x = mean_latitude, y = mean_elevation)) +
    geom_point(color = "blue") +
    geom_smooth(method = "lm", color = "red", se = TRUE) +
    labs(
      title = paste("Mean Elevation vs Latitude ", interval),
      x = "Mean Latitude (km North of Equator)",
      y = "Mean Elevation (m)"
    ) +
    theme_minimal()
  
  plots[[as.character(interval)]] <- p
}

grid.arrange(grobs = plots, ncol = 2)

```

Notes: In time from 1999 to 2010, the trend line shows a slight negative slope, indicating that bumblebee species at higher latitudes might be occupying lower elevations compared to those at lower latitudes.
This could reflect the influence of climate change, where higher latitude species face constraints on elevation ranges due to warming conditions.

## Plot 4: Change in Temp Vs Mean Latitude

```{r TempVariation}
library(ggplot2)
library(dplyr)

data <- data %>%
  mutate(temperature_variation = maxPeriodAnnualMeanT - minPeriodAnnualMeanT)

# Calculate the mean temperature variation and latitude for each species by time interval
species_temp_summary <- data %>%
  group_by(species, periodTo) %>%
  summarize(
    mean_temp_variation = mean(temperature_variation, na.rm = TRUE),
    mean_latitude = mean(kmNorthEquator, na.rm = TRUE)
  ) %>%
  ungroup()

ggplot(species_temp_summary, aes(x = mean_latitude, y = mean_temp_variation)) +
  geom_point(color = "blue") +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  facet_wrap(~periodTo, scales = "free") +  # Create one panel per time interval
  labs(
    title = "Mean Temperature Variation vs Latitude Across Time Intervals",
    x = "Mean Latitude (km North of Equator)",
    y = "Mean Temperature Variation (°C)"
  ) +
  theme_minimal()

```

Positive Correlation: In all time intervals, the red lines show a positive slope, suggesting that as the latitude increases (moving further north), the mean temperature variation tends to increase.
Strength of the Relationship:
The steepness of the regression line indicates the strength of the relationship. In some intervals (e.g., 1998, 2010), the slope is steeper, suggesting a stronger correlation between latitude and temperature variation.
Consistency Across Intervals: The positive relationship between latitude and temperature variation appears to be consistent across all intervals, but the magnitude of the variation differs.

## Plot 5: Mean Latitude Across time interval

```{r TimeInterval}
species_latitude_summary <- data %>%
  group_by(species, periodTo) %>%
  summarize(mean_latitude = mean(kmNorthEquator, na.rm = TRUE)) %>%
  ungroup()

# Plot latitude across time intervals
ggplot(species_latitude_summary, aes(x = as.factor(periodTo), y = mean_latitude)) +
  geom_boxplot(fill = "lightblue", color = "darkblue") +
  geom_jitter(width = 0.2, alpha = 0.5, color = "red") +
  labs(
    title = "Mean Latitude Across Time Intervals",
    x = "Time Interval",
    y = "Mean Latitude (km North of Equator)"
  ) +
  theme_minimal()
```

### Plot 6: Density/Histogram of Response

```{r plotnew}
library(ggplot2)

data <- data %>%
  mutate(tempVariation = maxPeriodAnnualMeanT - minPeriodAnnualMeanT)

hist(data$kmNorthEquator, main = "Histogram of Distance from Equator", xlab = "Distance from equator", probability = TRUE)

# Density plot
plot(density(data$kmNorthEquator), main = "Density Plot")
# Q-Q plot
qqnorm(data$kmNorthEquator)
qqline(data$kmNorthEquator, col = "red")

```