---
title: "LRegression406"
author: "Zhiling Huang"
date: "2024-11-30"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Monte Carlo for Overall Population

```{r overall}
library(dplyr)
library(ggplot2)
data <- read.csv("final_data_1.csv")

# Add temperature variation
data <- data %>%
  mutate(tempVariation = maxPeriodAnnualMeanT - minPeriodAnnualMeanT)

# Monte Carlo simulation function
monte_carlo_analysis <- function(data, n_iterations = 1000) {
  results <- replicate(n_iterations, {
    # Bootstrap sample
    sample_data <- data[sample(1:nrow(data), replace = TRUE), ]
    # Linear regression
    model <- lm(kmNorthEquator ~ tempVariation, data = sample_data)
    coef(model)
  })
  t(results)
}

# Perform Monte Carlo analysis on the entire dataset
overall_results <- monte_carlo_analysis(data)

# Summarize results
slopes1 <- overall_results[, "tempVariation"]
intercepts1 <- overall_results[, "(Intercept)"]

summary_list1 <- list(
  Mean_Slope = mean(slopes1),
  Mean_Intercept = mean(intercepts1),
  Lower_Bound_CI = quantile(slopes1, 0.025),
  Upper_Bound_CI = quantile(slopes1, 0.975),
  Reject_Null = !(0 >= quantile(slopes1, 0.025) & 0 <= quantile(slopes1, 0.975))
)
overall_summary_df <- as.data.frame(summary_list1)

print(overall_summary_df)

```
## Actual VS theortical

```{r plot}
library(ggplot2)
theoretical <- rnorm(length(slopes1), mean = mean(slopes1), sd = sd(slopes1))

# Actual and theoretical distributions
distribution_data <- data.frame(
  Value = c(slopes1, theoretical),
  Distribution = rep(c("Actual", "Theoretical"), each = length(slopes1))
)

ggplot(distribution_data, aes(x = Value, fill = Distribution)) +
  geom_density(alpha = 0.5) +
  labs(
    title = "Actual vs Theoretical Distribution of Slopes",
    x = "Slope Value (Temperature Variation)",
    y = "Density"
  ) +
  theme_minimal() +
  theme(legend.title = element_blank())
```
## Bootstrapped samples

```{r 11}
## Bootstrapped Samples
hist(slopes1, main = "Distribution of Bootstrapped Slopes", xlab = "Slope")

```

#### Break into time interval for overall bumblebee population

```{r 567}

time_periods <- list(
  c(1901, 1974),
  c(1975, 1986),
  c(1987, 1998),
  c(1999, 2010)
)

time_interval_result <- list()

for (period in time_periods) {
  period_data <- data %>%
    filter(periodFrom >= period[1] & periodTo <= period[2]) 
  
  if (nrow(period_data) > 0) {
    time_interval_result[[paste(period[1], period[2], sep = "-")]] <- monte_carlo_analysis(period_data)
  }
}

time_summary_list <- list()

for (period in names(time_interval_result)) {
  results <- time_interval_result[[period]]
  slopes <- results[, "tempVariation"]
  intercepts <- results[, "(Intercept)"]
  
  time_summary_list[[period]] <- list(
    Period = period,
    Mean_Slope = mean(slopes),
    Mean_Intercept = mean(intercepts),
    Lower_Bound_CI = quantile(slopes, 0.025),
    Upper_Bound_CI = quantile(slopes, 0.975),
    Reject_Null = !(0 >= quantile(slopes, 0.025) & 0 <= quantile(slopes, 0.975))
  )
}

new_summary_df <- do.call(rbind, lapply(time_summary_list, as.data.frame))

new_summary_df
```

## change of slope over time

```{r newplot}

time_summary_df <- do.call(rbind, lapply(time_summary_list, as.data.frame))
time_summary_df$Period <- factor(time_summary_df$Period, levels = c("1901-1974", "1975-1986", "1987-1998", "1999-2010"))

ggplot(time_summary_df, aes(x = Period, y = Mean_Slope, fill = Period)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7) +
  geom_errorbar(aes(ymin = Lower_Bound_CI, ymax = Upper_Bound_CI), width = 0.2) +
  labs(
    title = "Changes in Slope Across Time Periods",
    x = "Time Period",
    y = "Mean Slope (Temperature Variation vs Migration)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

### Linear Regression

Simple OLS

```{r 555}

model <- lm(kmNorthEquator ~ tempVariation + maxPeriodAnnualMeanT, data = data)
summary(model)
par(mfrow = c(2, 2)) 
plot(model)
```

#### Spline for Non-linearity

```{r p1}
library(splines)
model_spline <- lm(kmNorthEquator ~ ns(tempVariation, df = 3), data = data)
summary(model_spline)

par(mfrow = c(2, 2)) 
plot(model_spline)
```
Adding non-linear terms doesn't seems to improve overall r-squared values, in this case we can conclude a positive linear relationship between temperature and kmNorthequator should be as sufficient as non-linear coefficient.  